Precise spiking motifs in neurobiological and neuromorphic data

### 2022-11-15 manuscript.tex
5 reformulation

% retina
How may information about the sensory world be encoded in the spiking pattern?  At the level of the retina, it has been shown that a coding of luminance values in the image using the timing of the spikes may be at work~\citep{gollisch_rapid_2008}. In particular, these results show that the response of ganglion cells to the visual gratings projected on the retina could be encoded in the latency of the response and not only in the frequency of the discharge, as it is often assumed. These results have been extended to natural images and show a qualitatively similar behavior. The authors' conclusion is that the precise spiking latency of the neurons encodes the spatial features of the image. Interestingly, such a precise latency mechanism may underlie some visual illusions, e.g., the false color illusion in the \href{https://michaelbach.de/ot/col-Benham/index.html}{Benham Top} based on center-surround interactions in the parvocellular pathway~\citep{kenyon_theory_2004}. This evidence found in the retina can be extended to other areas such as the visual cortex.
%----------------------------%
% C_MainenSejnowski1995_Perrinet.ipynb
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/replicating_MainenSejnowski1995.pdf} 
\caption{\textbf{Reproducibility of the spiking response of a neuron. }The timing of the spikes produced following the repetition of a step stimulus is less reproducible than that to a noisy stimulus. The stimulus current value over time for a step stimulus (top left) and for a noisy one (top right). Trial repetitions of a Leaky Integrate-and-Fire neuron stimulated by the stimulus on the upper row (middle row). Membrane potential represented by dark blue color when light yellow colors when depolarized) and quantified by . The average firing rate across trials (lower row). While this seems paradoxical at first sight, this is due to the use of the same \emph{frozen} noise at each repetition and highlight the highly reproducible pattern of spikes when it is driven by a highly dynamic input. See this~\href{https://github.com/laurentperrinet/2022_UE-neurosciences-computationnelles/blob/master/C_MainenSejnowski1995_Perrinet.ipynb}{notebook} for a replication of this result using a simple LIF model.}\label{fig:mainen}
\end{figure}
%----------------------------%

@@

Theoretically, it was shown that a simple model may allow the propagation of such synfire chains~\citep{gewaltig_propagation_2001} by considering the dynamics of leaky integrate-and-fire (LIF) neurons in different groups of similar size. Each neuron of one group is connected by an excitatory synapse to the next. When a pulse is elicited in the first group, this may generate a spike in the next group. Depending on the concentration of synaptic weight values, this new activity may become more or less synchronized with respect to that of the previous pulse (as measured by the standard deviation of spike times within the pulse). Recursively applying this to a sequence of groups within a chain generates either a synfire propagation or not. A simple simulation of synfire propagation is shown in Figure~\ref{fig:diesman}. A crucial aspect of this emergence is explained by the dynamics of the spiking neuron model~\citep{gerstner_time_1995} and in particular the balance between excitation and inhibition~\citep{azouz_stimulus-selective_2008}. This balance was for instance modeled by feed-forward inhibition, a fine-scaled latency mechanism that is an essential ingredient in modelling so-called push-pull effects in the primary visual cortex~\citep{kremkow_push-pull_2016}. Further models have shown that such synfire chains could be embedded in topographies~\citep{aviel_embedding_2003}, while others used conductance-based neurons with feed-forward inhibition to improve the robustness of the propagation~\citep{kremkow_functional_2010}. In particular, this was implemented as a computational neuroscience benchmark model using the pyNN language~\citep{davison_pynn_2008} both in CPU-based and neuromorphic hardware~\citep{pfeil_six_2013}. % (see Figure~\ref{fig:pynn}).
%
%----------------------------%
%\begin{figure}
%\centering
%\includegraphics[width=.7\textwidth]{figures/fnins-07-00011-g004.jpg}
%\caption{A pyNN implementation of a Synfire chain with feedforward inhibition. The background noise  is implemented as random Gaussian current. The state space gives %the duration of synfire chains as a function of the protocol's parameters.}\label{fig:pynn}
%\end{figure}
%----------------------------%
%----------------------------%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/Diesmann_et_al_1999.pdf}
\caption{\textbf{Simulation of a synfire propagation \href{https://brian2.readthedocs.io/en/stable/examples/frompapers.Diesmann_et_al_1999.html}{using Brian}.} The model consists of $10$ groups (arranged with the first group represented in the lowest row) of $100$ neurons each. Five pulses with decreasing jitter are generated in the first group around times $10$, $30$, $50$, $70$ and $90$~\ms~(with jitters given by a standard deviation which linearly decreases from $5$ to $1$~\ms). This generates a pulse after a certain processing delay in the second group with a different jitter. While the first two pulses progressively vanish, from the third input pulse, it is propagated to the following groups. This allows the propagation of the synchronous activity along the chain of the neural groups.}\label{fig:diesman}
\end{figure}
%----------------------------%

@@

Additional studies detail the role of such precise spike timing in downstream information transfer and coding~\citep{villette_internally_2015,branco_dendritic_2010,luczak_packet-based_2015}. This is for instance relevant in sensory pathways in vision~\citep{meister_concerted_1995}, audition~\citep{decharms_primary_1996}, olfaction~\citep{wehr_odour_1996} or touch~\citep{johansson_first_2004}.  In particular, stereotyped sequences of neural activation have been described in the adult hippocampus and related to its function in mental travel in time and space~\citep{buzsaki_space_2018}. These sequences can be internally generated~\citep{pastalkova_internally_2008,villette_internally_2015} and may be formed by the chained activation of orthogonal assemblies, themselves organized as sequence packets~\citep{malvache_awake_2016}. In that protocol, hippocampal sequences are formed by the ordered activation of smaller sequence motifs. They are stereotyped and robust, since neurons can be activated in the same order across days (see Figure~\ref{fig:haimerl} from~\citep{haimerl_internal_2019}). As a consequence, hippocampal sequences may rely on an internally hardwired structure and form the functional building blocks for encoding, storing and retrieving experience.
%
%----------------------------%
\begin{figure}
\centering
\includegraphics[width=\linewidth]{figures/haimerl2019.jpg}
\caption{\textbf{An example of a precise temporal motifs seen in cortical slices.} In this study by~\citep{haimerl_internal_2019}, an analysis of the raster plot shows repetitions of precise spiking motifs with a time scale of the order of seconds. (A) Calcium fluorescence (heatmap) of CA1 neurons participating to run sequences in consecutive imaging sessions. Cells have been selected and ordered with respect to their activity in the first imaging session (Top). The black line represents the speed of the mouse (modified from Figure 1 from~\citep{haimerl_internal_2019} under the CC-BY \href{https://www.pnas.org/doi/full/10.1073/pnas.1718518116}{PNAS License}).}\label{fig:haimerl}
\end{figure}
%----------------------------%

@@

%----------------------------%
\begin{figure}
\centering
\includegraphics[width=\linewidth]{figures/pcbi.1006283.g001.png}%pcbi.1006283.g001.PNG_L.png}
\caption{\textbf{The SPOTDisClust algorithm.} Fig 1 of~\citep{grossberger_unsupervised_2018}, showing simulated example illustrating the steps in the algorihtm. \textbf{(A)} Structure of five ``ground-truth'' patterns (\ldots). For each pattern and each neuron, a random position was chosen for the activation pulse. \textbf{(B)} Neuronal output is generated according to an inhomogeneous Poisson process, with rates dictated by the patterns in (A). (© Authors under a \href{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1006283}{CC licence})}\label{fig:SPOTDisClust}
\end{figure}
%----------------------------%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Scaling up to very large scale data}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Over the past decade, tremendous technological advances across several disciplines have dramatically expanded the frontiers of experimentally accessible neuroscientific facts. Bridging across different spatial and temporal scales, combination of \textit{in vivo} two photon imaging, large population recording-array technologies, optogenetic circuit control tools, transgenic manipulations as well as large volume circuit reconstructions are now used to examine the function, structure and dynamics of neural networks on an unprecedented level of detail and precision. The daunting complexity of the biological reality revealed by these technologies highlights the importance of neurobiological knowledge to provide a conceptual bridge between abstract principles of brain function and their biological implementations within neural circuits. As a consequence, there is a growing need to scale these methods to larger amounts of data.

@@

A natural candidate to use these precise temporal patterns in the brain is to use Spiking Neural Networks (SNNs)~\citep{maass_networks_1997}. The approach which is at present most prominent in the SNN community is to use existing algorithms from machine learning and to adapt them to the specificity of spiking architectures~\citep{goltz_fast_2021}. One such example is to adapt the successes of deep learning algorithms and to transfer the back-propagation algorithm to SNNs, as it is the most widely used to tune the weights of a classical (non spiking) Neural Network. In a nutshell, it considers the system as implementing an input / output function and iteratively updates the weights according to the direction and magnitude of the error function gradient. Since the activation function of a spiking neuron is not differentiable, a recent popular approach consists in using a surrogate gradient~\citep{neftci_surrogate_2019} to ``cross-compile'' classical Neural Network to a spiking architecture~\citep{rueckauer_conversion_2017}. This approach is quite successful, and SNNs reach in some case a similar performance as their non spiking equivalent, for instance on the N-MNIST dataset for categorizing digits in a stream of events~\citep{susi_nmnsd-spiking_2021}. The main reason for adopting this approach instead of classical architectures is the possibility to benefit from dedicated neuromorphic low-energy hardware. However, most biological neural systems use spikes and are obviously more efficient than current state-of-the-art vision systems, both in terms of efficiency (accuracy), in speed (latency), and energy consumption. There is therefore an immense gap in the way we understand biology to translate it to the efficiency of SNNs.  To go beyond the state-of-the-art, we will focus here on one core computation of a spiking neuron, that is, is its ability to switch from the classical integrator mode (summing analog currents on its synapses) to a synchrony detector where it emits a spike whenever presynaptic spiking inputs are synchronized~\citep{abeles_role_1982,paugam-moisy_computing_2012}. To overcome the diversity of input presynaptic patterns, we will explore different existing architectures which are able to learn to detect stable ``polychronous'' events, that is, volleys of spikes which are stable up to certain onset time (see Figure~\ref{fig:THC}). These models will be compared in light of neuroscientific and computational perspectives. We review in this section theoretical and computational foundations of spiking motif detection. %In particular, our objective is to describe methods which fully exploit the capacity of spiking neurons to detect synchronous patterns.

@@

As we saw, most SNN and in particular those adapted from analogous deep-learning-like architectures, rely on an encoding of information based on a continuously varying firing-rate. A notable exception of a SNN using precise spike timings is the \textit{polychronization} model of~\citet{izhikevich_polychronization_2006}. This model is based on a random recurrent model of spiking neurons including synaptic delays chosen randomly from a range of biologically realistic delays (from $0$ to $20~\ms$) and whose weights evolved with a Spike-Time Dependent Plasticity (STDP) learning rule~\citep{markram_regulation_1997}. It is consensus that spike timing (STDP) plays a crucial role in the development of synaptic efficacy for many different kinds of neurons~\citep{caporale_spike_2008}. Delays are defined as the total time taken for a spike to be conducted from one presynaptic neuron's soma to the efferent postsynaptic neuron's soma. It is worth mentioning that only the weights are changed using the STDP rule and that the set of delays is set randomly at initialization and that delays are then ``frozen'' for the rest of the simulation. Due to the interplay between the delays and STDP, the spiking neurons spontaneously self-organize into groups and generate patterns of stereotypical polychronous activity, i.e.~exhibit reproducible time-locked firing patterns which  the author defined as ``polychronous groups'' (PGs). One core ingredient of this model is the fact that the neurons composing a group fire at different times, but due to the heterogeneous delays, the spikes reach the postsynaptic neuron at the same time. This synchrony of arrival at the soma of the neuron leads to the summation of the excitatory post-synaptic potentials evoked by each spike, and thus to the crossing of the voltage threshold and to the discharge of a spike (see Figure~\ref{fig:izhikevitch}). According to the STDP rule, the group of neurons involved in this polychronous activity will see their synaptic weight increase and, thus, may consolidate the formation of a polychronous group. %----------------------------%

@@

    \caption{\textbf{Core mechanism of polychrony detection~\citep{izhikevich_polychronization_2006}.} \textit{(Left)}~In this example, three presynaptic neurons denoted \textit{b}, \textit{c} and, \textit{d} are fully connected to two post-synaptic neurons \textit{a} and \textit{e}, with different delays of respectively $1$, $5$, and $9~\ms$ for \textit{a} and  $8$, $5$, and $1~\ms$ for \textit{e}. \textit{(Middle)}~If three synchronous pulses are emitted from presynaptic neurons, this will generate post-synaptic potentials that will reach \textit{a} and \textit{e} asynchronously because of the delays, and they may not be sufficient to reach the membrane threshold in either of the post-synaptic neurons, therefore no spike will be emitted as this is not sufficient to reach the membrane threshold of the post synaptic neuron, so no output spike is emitted.

@@

    \textit{(Right)}~If the pulses are emitted from presynaptic neurons such that, taking into account the delays, they reach the post-synaptic neuron \textit{a} at the same time (here, at $t=10~\ms$), then a spike is emitted (red color), while none is emitted from post-synaptic neuron \textit{e}.

@@

% follow-ups
Interestingly, the paper by~\citep{izhikevich_polychronization_2006} stirred a lively debate in the field of computational neuroscience, with a general positive acceptance, but relatively few works extended this seminal paper. Indeed, there were already existing models of synaptic delay learning in spiking neural networks, see for instance~\citep{huning_synaptic_1998} or~\citep{eurich_dynamics_1999} yet they had not shown potential applications to the detection of spiking motifs. A popular model for the detection of latency patterns is the \emph{tempotron}~\citep{gutig_tempotron_2006}, particularly reviewed in~\citep{gutig_spike_2014}. The \emph{tempotron} is a supervised synaptic learning algorithm which classifies a distractor from a target motif, in order to extend the perceptron, which does not incorporate a spike timing framework. The \emph{tempotron} learning rule is derived by an optimization process and takes the form of an STDP rule. The limits of this model are that its output is only binary and that its storage capacities are limited. An extension of~\citep{izhikevich_polychronization_2006} was made in a very detailed work aiming at reproducing the polychronization model~\citep{pauli_reproducing_2018}. Indeed, while the original paper contained material within the text to reproduce the whole model (using MATLAB), it was not complete such as to allow for the reproduction of all results presented in that manuscript. This more recent work details how this code could be slightly corrected. It comes with a Python code and a version control system detailing the whole process used to give provenance to the different steps in this scientific process. Another recent work gives a Bayesian account in a similar model~\citep{guise_bayesian_2014}. In that work, based on the fact that previous methods for studying polychronous groups' activation response to stimuli have been limited by the template-based methods used to identify PG activation, the authors outline a new method that overcomes these difficulties by establishing a probabilistic interpretation of PG activation. They demonstrate the use of this method by investigating the claim that PGs might provide the foundation of a representational system. Stimulation of the trained network produces the activation of a PG, i.e. the propagation of firing activity through multiple layers due to convergent patterns of firing. While extending the original method, these methods reveal shortcomings that we will try to analyze in the rest of this section.

Strikingly, thanks to the fact that a neuron can be involved in different polychronous groups, the number of coexisting polychronous groups far exceeds the number of neurons in the network, resulting in an unprecedented memory capacity of the system (see Figure~\ref{fig:THC}). In other neuronal models, an efficient use or detection of these spatio-temporal patterns embedded in the spike train comes with the integration of heterogeneous delays~\citep{guise_bayesian_2014,zhang_supervised_2020}. The recent ``multi-neuronal spike sequence detector'' architecture integrates the weight- and delay-adjustment methods by combining plasticity with the modulation of spike latency emission~\citep{susi_nmnsd-spiking_2021}. Additional models for detection of latency patterns are presented in the extensive (graph-centric) review on synchronization in time-varying networks~\citep{ghosh_synchronization_2021,ghosh_synchronized_2022}. This representation has potentially a much greater information capacity in comparison to other neural coding approaches through their connectivity and the possible coexistence of numerous superposed PGs~\citep{izhikevich_polychronous_2009}. Recently, by using a logistic regression model coupled with a temporal convolution implemented a model with heterogeneous delays was implemented to test the detection of the spiking motifs embedded in the event stream~\citep{grimaldi_learning_2022}. This allowed to detect the detection of a high number of superposed polychronous motifs in synthetic data, illustrating the computational benefit of such representations compared to that with a unique delay (see Figure~\ref{fig:THC}). As such, these models use the neural dynamics to handle input signals with different delays but do not explicitly take full advantage of the representation capacity offered by heterogeneous delays. 

@@

    \caption{\textbf{Detecting event-based motifs using spiking neurons with heterogeneous delays.} 

@@

At the implementation level, a recent work proposed a bio-plausible unsupervised delay learning for extracting temporal features in spiking neural networks~\citep{nadafian_bio-plausible_2020}. The authors provided some mathematical proofs to show that their learning rule gives the ability to learn repeating spatio-temporal patterns. Applying this STDP-based rule on delays to the spiking neural network, experimental results were validated on a simple motion detection task, but were prone to convergence issues. Another model of synaptic delay-weight plasticity integrates synaptic delay plasticity into supervised learning and proposes a novel learning method that adjusts both the synaptic delays and weights of the learning neurons to make them fire precisely timed spikes~\citep{zhang_supervised_2020}. This was also presented by~\citep{wang_delay_2019} which propose a supervised delay learning algorithm for spiking neurons with temporal encoding, in which both the weight and delay of a synaptic connection can be adjusted to enhance the learning performance. Other models like that of~\citep{hazan_memory_2022} propose a weightless spiking neural networks that can perform a simple classification task which is applied to MNIST. In a recent paper~\citep{luo_supervised_2022}, authors propose a gradient descent-based learning algorithm for synaptic delays to enhance the sequential learning performance of a single spiking neuron. In this algorithm, information is encoded in the relative timing of individual neuronal spikes, and learning is performed based on the exact derivatives of the postsynaptic spike times with respect to presynaptic spike times.
In yet another computational model,~\citet{sun_learning_2016} show that the frequently activated polychronous neural groups can be learned efficiently by readout neurons with joint weight-delay spike-timing-dependent plasticity.

>>>>

% retina
How may information about the sensory world be encoded in the spiking pattern?  At the level of the retina, it has been shown that a coding of luminance values in the image using the timing of the spikes may be at work~\citep{gollisch_rapid_2008}. In particular, these results show that the response of ganglion cells to the visual gratings projected on the retina could be encoded in the latency of the response and not only in the frequency of the discharge, as it is often assumed. These results have been extended to natural images and show a qualitatively similar behavior. The authors' conclusion is that the precise spiking latency of the neurons encodes the spatial features of the image. Interestingly, such a precise latency mechanism may underlie some visual illusions, e.g., the false color illusion in the \href{https://michaelbach.de/ot/col-Benham/index.html}{Benham Top} based on center-surround interactions in the parvocellular pathway~\citep{kenyon_theory_2004}. This evidence found in the retina can be extended to other areas such as the visual cortex.
%----------------------------%
% C_MainenSejnowski1995_Perrinet.ipynb
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/replicating_MainenSejnowski1995.pdf} 
\caption{The timing of the spikes produced following the repetition of a step stimulus is less reproducible than that to a noisy stimulus. The stimulus current value over time for a step stimulus (top left) and for a noisy one (top right). Trial repetitions of a Leaky Integrate-and-Fire neuron stimulated by the stimulus on the upper row (middle row). Membrane potential represented by dark blue color when light yellow colors when depolarized) and quantified by . The average firing rate across trials (lower row). While this seems paradoxical at first sight, this is due to the use of the same \emph{frozen} noise at each repetition and highlight the highly reproducible pattern of spikes when it is driven by a highly dynamic input. See this~\href{https://github.com/laurentperrinet/2022_UE-neurosciences-computationnelles/blob/master/C_MainenSejnowski1995_Perrinet.ipynb}{notebook} for a replication of this result using a simple LIF model.}\label{fig:mainen}
\end{figure}
%----------------------------%

@@

Theoretically, it was shown that a simple model may allow the propagation of such synfire chains~\citep{gewaltig_propagation_2001} by considering the dynamics of leaky integrate-and-fire (LIF) neurons in different groups of similar size. Each neuron of one group is connected by an excitatory synapse to the next. When a pulse is elicited in the first group, this may generate a spike in the next group. Depending on the concentration of synaptic weight values, this new activity may become more or less synchronized with respect to that of the previous pulse (as measured by the standard deviation of spike times within the pulse). Recursively applying this to a sequence of groups within a chain generates either a synfire propagation or not. A simple simulation of synfire propagation is shown in Figure~\ref{fig:diesman}. A crucial aspect of this emergence is explained by the dynamics of the spiking neuron model~\citep{gerstner_time_1995} and in particular the balance between excitation and inhibition~\citep{azouz_stimulus-selective_2008}. This balance was for instance modeled by feed-forward inhibition, a fine-scaled latency mechanism that is an essential ingredient in modelling so-called push-pull effects in the primary visual cortex~\citep{kremkow_push-pull_2016}. Further models have shown that such synfire chains could be embedded in topographies~\citep{aviel_embedding_2003}, while others used conductance-based neurons with feed-forward inhibition to improve the robustness of the propagation~\citep{kremkow_functional_2010}. In particular, this was implemented as a computational neuroscience benchmark model using the pyNN language~\citep{davison_pynn_2008} both in CPU-based and neuromorphic hardware~\citep{pfeil_six_2013}. % (see Figure~\ref{fig:pynn}).
%
%----------------------------%
%\begin{figure}
%\centering
%\includegraphics[width=.7\textwidth]{figures/fnins-07-00011-g004.jpg}
%\caption{A pyNN implementation of a Synfire chain with feedforward inhibition. The background noise  is implemented as random Gaussian current. The state space gives %the duration of synfire chains as a function of the protocol's parameters.}\label{fig:pynn}
%\end{figure}
%----------------------------%
%----------------------------%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/Diesmann_et_al_1999.pdf}
\caption{Simulation of a synfire propagation \href{https://brian2.readthedocs.io/en/stable/examples/frompapers.Diesmann_et_al_1999.html}{using Brian}. The model consists of $10$ groups (arranged with the first group represented in the lowest row) of $100$ neurons each. Five pulses with decreasing jitter are generated in the first group around times $10$, $30$, $50$, $70$ and $90$~\ms~(with jitters given by a standard deviation which linearly decreases from $5$ to $1$~\ms). This generates a pulse after a certain processing delay in the second group with a different jitter. While the first two pulses progressively vanish, from the third input pulse, it is propagated to the following groups. This allows the propagation of the synchronous activity along the chain of the neural groups.}\label{fig:diesman}
\end{figure}
%----------------------------%

@@

Additional studies detail the role of such precise spike timing in downstream information transfer and coding~\citep{villette_internally_2015,branco_dendritic_2010,luczak_packet-based_2015}. This is for instance relevant in sensory pathways in vision~\citep{meister_concerted_1995}, audition~\citep{decharms_primary_1996}, olfaction~\citep{wehr_odour_1996} or touch~\citep{johansson_first_2004}.  In particular, stereotyped sequences of neural activation have been described in the adult hippocampus and related to its function in mental travel in time and space~\citep{buzsaki_space_2018}. These sequences can be internally generated~\citep{pastalkova_internally_2008,villette_internally_2015} and may be formed by the chained activation of orthogonal assemblies, themselves organized as sequence packets~\citep{malvache_awake_2016}. In that protocol, hippocampal sequences are formed by the ordered activation of smaller sequence motifs. They are stereotyped and robust, since neurons can be activated in the same order across days (see Figure~\ref{fig:haimerl} from~\citep{haimerl_internal_2019}). As a consequence, hippocampal sequences may rely on an internally hardwired structure and form the functional building blocks for encoding, storing and retrieving experience.
%
%----------------------------%
\begin{figure}
\centering
\includegraphics[width=\linewidth]{figures/haimerl2019.jpg}
\caption{An example of a precise temporal motifs seen in cortical slices. In this study by~\citep{haimerl_internal_2019}, an analysis of the raster plot shows repetitions of precise spiking motifs with a time scale of the order of seconds. (A) Calcium fluorescence (heatmap) of CA1 neurons participating to run sequences in consecutive imaging sessions. Cells have been selected and ordered with respect to their activity in the first imaging session (Top). The black line represents the speed of the mouse (modified from Figure 1 from~\citep{haimerl_internal_2019} under the CC-BY \href{https://www.pnas.org/doi/full/10.1073/pnas.1718518116}{PNAS License}).}\label{fig:haimerl}
\end{figure}
%----------------------------%

@@

%----------------------------%
\begin{figure}
\centering
\includegraphics[width=\linewidth]{figures/pcbi.1006283.g001.png}%pcbi.1006283.g001.PNG_L.png}
\caption{Fig 1 of~\citep{grossberger_unsupervised_2018}, showing simulated example illustrating the steps in SPOTDisClust. A) Structure of five ``ground-truth'' patterns (\ldots). For each pattern and each neuron, a random position was chosen for the activation pulse. B) Neuronal output is generated according to an inhomogeneous Poisson process, with rates dictated by the patterns in (A). (© Authors under a \href{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1006283}{CC licence})}\label{fig:SPOTDisClust}
\end{figure}
%----------------------------%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Scaling up to very large scale data}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Over the past decade, tremendous technological advances across several disciplines have dramatically expanded the frontiers of experimentally accessible neuroscientific facts. Bridging across different spatial and temporal scales, combination of \textit{in vivo} two photon imaging, large population recording-array technologies, optogenetic circuit control tools, transgenic manipulations as well as large volume circuit reconstructions are now used to examine the function, structure and dynamics of neural networks on an unprecedented level of detail and precision. The daunting complexity of the biological reality revealed by these technologies highlights the importance of neurobiological knowledge to provide a conceptual bridge between abstract principles of brain function and their biological implementations within neural circuits. As a consequence, there is a growing need to scale these methods to larger amounts of data.

@@

A natural candidate to use these precise temporal patterns in the brain is to use Spiking Neural Networks (SNNs)~\citep{maass_networks_1997}. The approach which is currently most prominent in the SNNs community is to use existing algorithms from machine learning and to adapt them to the specificity of spiking architectures~\citep{goltz_fast_2021}. One such example is to adapt the successes of deep learning algorithms and to transfer the back-propagation algorithm to SNNs. This algorithms is the most widely used algorithm to tune the weights of a classical (non spiking) Neural Network. It iteratively updates the weights according to the direction and magnitude of the error function gradient. Since the activation function of a spiking neuron is not differentiable, a recent popular approach consists in using a surrogate gradient~\citep{neftci_surrogate_2019} to ``cross-compile'' classical Neural Network to a spiking architecture~\citep{rueckauer_conversion_2017}. This approach is quite successful, and SNNs reach in some case a similar performance as their non spiking equivalent, for instance on the N-MNIST dataset for categorizing digits in a stream of events~\citep{susi_nmnsd-spiking_2021}. The main reason for adopting this approach is to benefit from dedicated neuromorphic low-energy hardware. However, most biological neural systems use spikes and are obviously more efficient than current state-of-the-art vision systems, both in terms of efficiency (accuracy), in speed (latency), and energy consumption. There is therefore an immense gap in the way we understand biology to translate it to the efficiency of SNNs.  To go beyond the state-of-the-art, we will focus here on one core computation of a spiking neuron, that is, is its ability to switch from the classical integrator mode (summing analog currents on its synapses) to a synchrony detector where it emits a spike whenever presynaptic spiking inputs are synchronized~\citep{abeles_role_1982,paugam-moisy_computing_2012}. To overcome the diversity of input presynaptic patterns, we will explore different existing architectures which are able to learn to detect stable ``polychronous'' events, that is, volleys of spikes which are stable up to certain synaptic delays. These models will be compared in light of neuroscientific and computational perspectives. We review in this section theoretical and computational foundations of PG detection in models. In particular, our objective is to describe methods which fully exploit the capacity of spiking neurons to detect synchronous patterns.

@@

As we saw, most SNN and in particular those adapted from analogous deep-learning-like architectures, rely on an encoding of information based on a continuously varying firing-rate. A notable example of a SNN using precise spike timings is the polychronization model of~\citet{izhikevich_polychronization_2006}. This model is based on a random recurrent model of spiking neurons including synaptic delays chosen randomly from a range of biologically realistic delays (from $0$ to $20~\ms$) and whose weights evolved with a Spike-Time Dependent Plasticity (STDP) learning rule~\citep{markram_regulation_1997}. It is consensus that spike timing (STDP) plays a crucial role in the development of synaptic efficacy for many different kinds of neurons~\citep{caporale_spike_2008}. It is worth mentioning that only the weights are changed using the STDP rule and that the set of delays is set randomly at initialization and that it is then frozen for the rest of the simulation. Delays are defined as the total time taken for a spike to be conducted from one presynaptic neuron's soma to the efferent postsynaptic neuron's soma. Due to the interplay between the delays and STDP, the spiking neurons spontaneously self-organize into groups and generate patterns of stereotypical polychronous activity, i.e.~exhibit reproducible time-locked firing patterns. On core ingrredient of this model is the fact that the neurons composing a group fire at different times, but due to the heterogeneous delays, the spikes reach the postsynaptic neuron at the same time. This synchrony of arrival at the soma of the neuron leads to the summation of the excitatory post-synaptic potentials evoked by each spike, and thus to the crossing of the voltage threshold and to the discharge of a spike (see Figure~\ref{fig:izhikevitch}). According to the STDP rule, the neurons involved in this activity will see their synaptic weight increase and, thus, form a polychronous group. %----------------------------%

@@

    \caption{\textbf{Core mechanism of polychrony detection~\citep{izhikevich_polychronization_2006}.} \textit{(Left)}~In this example, three presynaptic neurons denoted \textit{b}, \textit{c} and, \textit{d} are fully connected to two post-synaptic neurons \textit{a} and \textit{e}, with different delays of respectively $1$, $5$, and $9~\ms$ for \textit{a} and  $8$, $5$, and $1~\ms$ for \textit{e}. \textit{(Middle)}~If three synchronous pulses are emitted from presynaptic neurons, this will generate post-synaptic potentials that will reach \textit{a} and \textit{e} asynchronously because of the delays, and they may not be sufficient to reach the membrane threshold in either of the post-synaptic neurons, therefore no spike will be emitted.

@@

    \textit{(Right)}~If the pulses are emitted from presynaptic neurons such that, taking into account the delays, these converge at the same time on neuron \textit{a} (here, at $t=10~\ms$) then a spike is emitted (here in red color), while none is emitted on neuron \textit{e}.

@@

% follow-ups
Interestingly, the paper by~\citep{izhikevich_polychronization_2006} stirred a lively debate in the field of computational neuroscience, with a general positive acceptance, but relatively few works extended this seminal paper. Indeed, there were already existing models of synaptic delay learning in spiking neural networks, see for instance~\citep{huning_synaptic_1998} or~\citep{eurich_dynamics_1999} yet they had not shown potential applications to the detection of spiking motifs. A popular model for the detection of latency patterns is the \emph{tempotron}~\citep{gutig_tempotron_2006}, particularly reviewed in~\citep{gutig_spike_2014}. The \emph{tempotron} is a supervised synaptic learning algorithm which classifies a distractor from a target motif, in order to extend the perceptron, which does not incorporate a spike timing framework. The \emph{tempotron} learning rule is derived by an optimization process and takes the form of an STDP rule. The limits of this model are that its output is only binary and that its storage capacities are limited. An extension of~\citep{izhikevich_polychronization_2006} was made in a very detailed work aiming at reproducing the polychronization model~\citep{pauli_reproducing_2018}. Indeed, while the original paper contained material within the text to reproduce the whole model (using MATLAB), it was not complete for the reproduction of all results presented in that manuscript. This more recent work details how this code was slightly corrected. It comes with a Python code and a version control system detailing the whole process used to give provenance to the different steps in this scientific process. Another recent work gives a Bayesian account in a similar model~\citep{guise_bayesian_2014}. In that work, based on the fact that previous methods for studying polychronous groups' activation response to stimuli have been limited by the template-based methods used to identify PG activation, the authors outline a new method that overcomes these difficulties by establishing a probabilistic interpretation of PG activation. They demonstrate the use of this method by investigating the claim that PGs might provide the foundation of a representational system. Stimulation of the trained network produces the activation of a PG, i.e. the propagation of firing activity through multiple layers due to convergent patterns of firing. While extending the original method, these methods reveal shortcomings that we will try to analyze in the rest of this section.

Interestingly, thanks to the fact that a neuron can be involved in different polychronous groups, the number of coexisting polychronous groups far exceeds the number of neurons in the network, resulting in an unprecedented memory capacity of the system (see Figure~\ref{fig:THC}). In other neuronal models, an efficient use or detection of these spatio-temporal patterns embedded in the spike train comes with the integration of heterogeneous delays~\citep{guise_bayesian_2014,zhang_supervised_2020}. The recent ``multi-neuronal spike sequence detector'' architecture integrates the weight- and delay-adjustment methods by combining plasticity with the modulation of spike latency emission~\citep{susi_nmnsd-spiking_2021}. Additional models for detection of latency patterns are presented in the extensive (graph-centric) review on synchronization in time-varying networks~\citep{ghosh_synchronization_2021,ghosh_synchronized_2022}. This representation has potentially a much greater information capacity in comparison to other neural coding approaches through their connectivity and the possible coexistence of numerous superposed PGs~\citep{izhikevich_polychronous_2009}. Recently, by using a logistic regression model coupled with a temporal convolution implemented a model with heterogeneous delays was implemented to test the detection of the spiking motifs embedded in the event stream~\citep{grimaldi_learning_2022}. This allowed to detect the detection of a high number of superposed polychronous motifs in synthetic data, illustrating the computational benefit of such representations compared to that with a unique delay (see Figure~\ref{fig:THC}). As such, these models use the neural dynamics to handle input signals with different delays but do not explicitly take full advantage of the representation capacity offered by heterogeneous delays. 

@@

    \caption{Detecting event-based motifs using spiking neurons with heterogeneous delays. 

@@

At the implementation level, a recent work proposed a bio-plausible unsupervised delay learning for extracting temporal features in spiking neural networks~\citep{nadafian_bio-plausible_2020}. The authors provided some mathematical proofs to show that their learning rule gives the ability to learn repeating spatio-temporal patterns. Applying this STDP-based rule on delays to the spiking neural network, experimental results were validated on a simple motion detection task, but were prone to convergence issues. Another model of synaptic delay-weight plasticity~\citep{zhang_supervised_2020} integrates synaptic delay plasticity into supervised learning and proposes a novel learning method that adjusts both the synaptic delays and weights of the learning neurons to make them fire precisely timed spikes. This was also presented by~\citep{wang_delay_2019} which propose a supervised delay learning algorithm for spiking neurons with temporal encoding, in which both the weight and delay of a synaptic connection can be adjusted to enhance the learning performance. Other models like that of~\citep{hazan_memory_2022} propose a weightless spiking neural networks that can perform a simple classification task which is applied to MNIST. In a recent paper~\citep{luo_supervised_2022}, authors propose a gradient descent-based learning algorithm for synaptic delays to enhance the sequential learning performance of a single spiking neuron. In this algorithm, information is encoded in the relative timing of individual neuronal spikes, and learning is performed based on the exact derivatives of the postsynaptic spike times with respect to presynaptic spike times.
In yet another computational model,~\citet{sun_learning_2016} show that the frequently activated polychronous neural groups can be learned efficiently by readout neurons with joint weight-delay spike-timing-dependent plasticity.
